---
title: "Lab 06 - Ugly charts and Simpson's paradox"
author: "Alex Connolly"
date: "Mar 10 2022"
output: 
  html_document: 
    fig_width: 9
    fig_height: 7
    highlight: haddock
---

### Load packages and data

```{r load-packages, message = FALSE}
library(tidyverse) 
library(dsbox)
library(mosaicData) 
```

```{r read}
staff <- read_csv("data/instructional-staff.csv")
```

```{r long}
staff_long <- staff %>%
  pivot_longer(cols = -faculty_type, names_to = "year") %>%
  mutate(value = as.numeric(value))
staff_long
```

```{r graph}
staff_long %>%
  ggplot(aes(x = year, y = value, color = faculty_type)) +
  geom_line()
```

### Exercise 1

```{r bigger graph}
staff_long %>%
  ggplot(aes(x = year, y = value, group = faculty_type, color = faculty_type)) + labs(title= "Trends in Instructional Staff Employees", color = "Type of Faculty", x= "Year", y= "Percentage of Hires" ) + 
  geom_line() 
```

### Exercise 2

You could also combine the other positions and do a part time vs everything else

### Exercise 3

```{r fish}
fisheries <- read_csv("data/fisheries.csv")
```





```{r pivot}
fisheries_new <- fisheries %>%
  pivot_longer(cols = -country, names_to = "Harvest") %>%
  filter(Harvest %in% c("capture", "aquaculture")) %>%
  mutate(value = as.numeric(value))
fisheries_new
```
```{r new}
fisheries_new <- fisheries_new %>%
  filter(country %in% c("China", "Russia", "Norway", "Malaysia", "Spain", "Nigeria", "France", "Mauritania", "Ghana", "Mozambique", "Poland", "Sierra Leone", "Laos", "Panama", "Seychelles", "Saudi Arabia"))
```
```{r graphh}
fisheries_new %>%
  ggplot(mapping = aes(fill = Harvest, y=value, x=country)) + geom_bar(position="stack", stat="identity") + scale_y_log10()
```


I created a new variable, harvest, putting both aquaculture and capture in it so I could create a stacked bar chart. I then log scaled the y axis so you could see more. I then kept only the countries with the most, the ones that were in the chart provided. This made it look much better. I could also adjust the x axis to make it easier to read. 


### Exerciser 4

```{r 2}
library(tidyverse) 
library(mosaicData) 
?Whickham
```

This is likely observational data since it is a survey.

```{r observe}
nrow(Whickham)
ncol(Whickham)
```
1314 observations, each is a person.
3 variables, representing if they are alive/dead, smoker or not, and age.
I would expect the smokers to be more likely to be less healthy. 

```{r whick}
Whickham %>%
  ggplot(mapping = aes(fill = outcome, x=smoker)) + geom_bar()
```

```{r}
Whickham %>%
  count(smoker, outcome)
```
68.6% of Non smokers are alive
76% of smokers are alive

```{r mutate}
Whickham <- Whickham %>%
  mutate(age_cat = case_when(
    age <= 44 ~ "18-44",
    age > 44 & age <= 64 ~ "45-64",
    age > 64 ~ "65+"
))
```
```{r facet}
Whickham %>%
  ggplot(mapping = aes(fill = outcome, x=smoker)) + facet_wrap(~age_cat)+geom_bar() 
```

```{r age}
Whickham %>%
  count(smoker, age_cat, outcome)
```
You can see that at a young age, smoking doesnt imapct health status. In middle age, the smokers start to die more, and in old age is where smoking really shows its head. 
